An anthropological account of the\\ Vim text editor:\\ features and tweaks\\ after 10 years of ouroboros

### 2018-02-11 article/article.tex
atualizando para repasse

\subsection{Further remarks about this document}
This document is written is a DRY KISS
(Don't Repeat Yourself, Keep It Simple Stupid) style.
Complex is to master the use of Vim
and one finds sound
references in help files and a nice vimrc.
Therefore the following content is kept
as uncomplicated and original as possible.
Also, because of Vim's complexity and entailed
bond of this document to my usage,
there is an anthropological component
which is evident in the occasional use of the first person.
This can be understood as anthropological computer science~\cite{anPh,anPh2}
and considered to help
in the technological groundwork of the civil society.
Accordingly, here are some notes about my experience:
I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
I used Vim for writing and editing computer code 
(Python, Javascript, C++, ChucK, bash, etc), markup-like content
(HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
Eventually, I edited other types of files, such as database files with
dozens of gigabytes.
Within Vim, I mostly write (web and scientific) software,
scientific articles, music, poems and short stories.

@@

\subsection{The bare minimum}\label{minimum}
You open a file at Vim startup by executing
the command: \texttt{vim <filename>}.
Inside Vim, you start in the normal
mode, and might want to move around using
\texttt{h-j-k-l} for left-down-up-right.
To insert characters, move your
cursor to the desired location an press i,
which puts Vim in the insert mode.
Go back to normal mode by pressing
\texttt{<ESC>}, \ttt{C-$[$} or \texttt{<C-C>}.
You save the file by typing \texttt{:w<CR>},
and exit Vim by typing \texttt{:q<CR>}.
You can save and quit with \tttt{:wq<CR>}
or \tttt{:x<CR>} or \tttt{ZZ}.

\subsection{Vim help}
Help on using Vim is found in various places.
The standard resource is the Vim help files.
They are accessed by typing \texttt{:h <anything><CR>}
in normal mode.
Examples of such \texttt{<anything>} are:
color, navigation, :vs, vimtutor.
Type 
\texttt{:h usr\_toc<CR>}
to access the official User Manual,
which is considerably lengthy and complex
and is usually not read by users before a few years
of experience.
In learning Vim, one
might want to run the \texttt{vimtutor} command
(outside Vim) to start the Vim Tutor.

There are good resources on the Web for learning
and tweaking Vim:
\begin{itemize}
  \item ``Vim Adventures'' is an online RPG game for practicing
  and memorizing Vim commands. This game is quite famous among Vim users.
  \item There are official and semi-official Vim sites e.g.:
  \url{www.vim.org}, \url{https://www.vi-improved.org} and
  \url{http://vim.wikia.com/}.
  \item Many hacks, understandings and general issues
  (e.g. how to make such a move) are asked and answered
  in online platforms (e.g. Quora, Stack Overflow, Stack Exchange, Reddit, Email list, IRC Channel).
  One often finds these links through a search engine.
  \item Many videos about Vim are publicly available.
  One traditional site is \url{http://vimcasts.org},
  but you might find them using a search engine
  (e.g. \url{http://derekwyatt.org/vim/tutorials/}) or in Youtube and Vimeo.
\end{itemize}

% any other commands? files?
\subsection{Namespaces}\label{namespaces}
Vim is a text editor ouroboros~\cite{ouroWiki} because
text and writing alters text and writing.
What you end with is a collection of namespaces where tokens have scalar
or arbitrarily complex values.

\subsubsection{Commands and mappings}
There are commands, i.e. typing sequences which trigger automated actions,
for each mode:
\begin{itemize}
	\item in Normal mode all keys are mapped to commands.
		There is redundancy and additional commands
		using Ctrl and Shift keys.
		Some keys expect a second key,
    and have combinations not used (thus available for new mappings),
		specially the z and g.
    See Section~\ref{normal} and Appendix~\ref{notes}
    for more insights into the commands available in the
    normal mode.
	\item In the other modes, the sequences available for mappings are more obvious and abundant.
		One should look at \tttt{:h index} to know about all the standard mappings
		and use \tttt{:map} to list the user-defined mappings.
\end{itemize}

@@

A colon command can be written as a string
and executed by the \tttt{:execute} colon command.
E.g. \tttt{:execute 'vs afile.txt'}.
As there are colon commands that execute commands in other
modes, e.g. \tttt{:normal ?\textasciicircum def},
the \tttt{:execute} is a way to build commands in any mode,
e.g.
\\\tttt{:execute 'normal i' . string(atan(bufnr('\%')))}.

\subsubsection{Variables}
There are some types of variables in Vim:
\begin{itemize}
	\item Environment variables: names start with \$ and hold system
		variables, such as \tttt{\$PATH} and \tttt{\$PWD}.
	\item Option variables: names start with \& and are meant to control the behavior of the editor.
		One might change a value through set or let, e.g.
		\tttt{:set bg=light} or \tttt{:let \&bg=light}.
	\item Registers: start with @ and are meant for automation and transfer of texts (copy and paste).
    More on register in Section~\ref{state} and scattered in this
    document.
	\item Internal variables are created with let and preferably have a prefix:
	\tttt{b:}, \tttt{w:}, \tttt{t:}, \tttt{l:}, \tttt{s:},
		are local to the buffer, window, tab page, function, and
		sourced Vim file, respectively.
 \tttt{v:}, \tttt{g:} are global, the first are predefined by Vim.
		\tttt{a:} is for function arguments.
		If there is no prefix, the variable is global or internal to a function if occurring inside a function.
		More about internal variables in \tttt{:h internal-variables}.
	\item The value of a variable can be a scalar, string, list, dictionary, function reference, etc (see \tttt{:h eval}).
\end{itemize}
You can echo any of such variables or use them in expressions.
Notice that you will only be able to echo a \tttt{b:} variable inside
the buffer where it is defined.
For all the Vimscript capabilities, including loops, conditionals,
and builtin functions, refer to \tttt{:h vim-script} and Section~\ref{script}.
Classes are possible only in rudimentary forms, e.g. through dictionaries,
but the language is otherwise overall quite powerful,
specially in dealing with text and editor behavior, as expected.

\subsubsection{State lists}\label{state}
Vim keeps a number of useful lists which expresses the state of the editor:
\begin{itemize}
	\item All the entered commands are accessed through \tttt{:hist a}. 
    The tokens ``\tttt{a / e : i d}'' may be used for specific types of commands, such
		as search and colon commands.
	\item File buffers are kept with numeric ids. See buffers with \tttt{:ls} and load a buffer to the window with \tttt{:b <num or token in file name>}.
  \item The windows open are listed in \tttt{:ls} with a character \tttt{a} in the second column, and are listed with \tttt{:tabs}.
	\item Tabs list can be reached through \tttt{:tabs}.
		It is usual both to show and hide the tabs bar (mapping
    in~\cite{vimrc}, discussion in Section~\ref{visual1}).
	\item Jumps are available through \tttt{:jumps}.
		One positions the cursor at each jump through \ttt{C-O} and \ttt{C-I}.
  \item Registers are available through the \tttt{:reg} command, as variables and through shortcuts in different modes.
    They also keep track or your copy, edition and deletion and are promptly defined
    by recording a typing sequence with the \tttt{q} normal command.
    Vim keeps only the last edition, in register \tttt{".}.
    An autocommand to keep the four latest inserts is in~\cite{vimrc}.
    A hack to keep the latest deletions and copies in the standard register
    might follow the same pattern, but use another \tttt{:h event} and
    monitor register \tttt{""} (maybe also monitor \tttt{"0}).
	\item An undo list can be accessed through \tttt{:changes}.
	\item A list with all the sourced scripts in a Vim instance is displayed through \tttt{:scriptnames}.
	\item The markers defined are listed with the \tttt{:marks} command.
		These are set by \tttt{mX} in normal mode, where X is the marker identifier.
		Uppercase letters are cross buffers.
	\item Quickfix and Location lists which are populated through \tttt{:vim} and \tttt{:make}
		and variations, such as \tttt{:grep}.
		One might run \tttt{:vim /section/ \%} and then \tttt{:copen}
		to open the Quickfix window, where the lines of occurrence are in sequence
		and one can \ttt{CR} one of them to have the cursor in the main window active at
		the first character of the match.
		One might run \tttt{:lvim /section/ \%} and then \tttt{:lopen}
		to use the location-list window instead of the Quickfix, which
		is very similar, but one per window instead of one per buffer.
		More information in \tttt{:h quickfix}.
	\item A tags list have to be made so one can use tags.
		Most often one will generate the tags list using the exuberant ctags,
		which supports dozens of languages.
		E.g. \tttt{:!ctags-exuberant functions.py} or \tttt{!ctags-exuberant -R ./},
		and then using \ttt{C-]} to
		go to the position of tag under cursor, and \tttt{:vs tags} to open
		the tags file.
	\item The argument list holds a list of files to be edited or browsed. 
		The list can be input at Vim startup (e.g. \tttt{\$ vim file1.txt file2.py})
	or using commands (e.g. \tttt{:ar ./*}).
		The file being edited is changed by \tttt{:n} and \tttt{:p} commands,
		one might perform actions on each file in argument list using \tttt{:argdo}.
		All files in argument list are also in the buffers list.
    For further information, see \tttt{:h arglist} and Section~\ref{netrw}.
  \item All the autocommands might be accessed through \tttt{:au}.
    They are event-triggered actions Vim performs.
\end{itemize}

A file with information about the state of the editor
can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim}.
Also, this script might be examined because it has
a collection of commands to access various settings of Vim.
Another good list of commands to know about Vim's state is kept on
\url{http://vim.wikia.com/wiki/Displaying_the_current_Vim_environment}.
I would specially highlight the \tttt{:syntax} command because
it displays token groups and their meanings when run
inside e.g. a \tttt{.py}, \tttt{.vim} or help file.

\subsubsection{On the persistence of visual cues about the editor state}\label{visual1}
You can keep track of the editor state though commands, as stated
above.
Also, one might rely on persistent visual
cues, specially the tabs bar, the status line, and the line
reserved for the command-line.
A good strategy I find is to have selective visual cues of the state to make persistent or hide
and a mapping to toggle each of them.
Currently, I toggle byobu/screen/tmux bar with \ttt{F5},
status line and tabs bar with \ttt{localleader-T or B} according to script~\cite{vimrc}.
I am mostly using the cleanest setting, toggling on the tabs bar and status line sometimes.
Numbering is always there, I rarely turn them off but keep the mappings \ttt{leader-n or N} to toggle just in case.
Instead of keeping the status bar, I use \ttt{C-G} to know about the
file and \ttt{gC-G} to know more and rarely.
It seems not possible to remove the statusbar between horizontal splits.
After asking in online forums and experimenting, I realized that it seems
reasonable to keep at least one line dividing the windows, so
if it comes to it, I just \tttt{set statusline=-}.
Unfortunately, as far as I could dig, one will need to alter
the Vim's source code to enable a horizontal split without losing a line.
For me, it would be ideal for this feature to have a visual cue of the
first and/or last line of the windows in the lines-number
column, or complete the spaces and empty chars with \$\$\$\$ or so.

%%%%%%%%%%%% TTM
\subsubsection{On the persistence of the editor state}
For state persistence, one might keep an undo file for each file as
in~\cite{vimrc}.
Sessions are easy to manage, enabling one to save and load the
editor's state, with the opened windows, tabs, buffers, etc.
The mappings in~\cite{vimrc} keep the sessions
in a reasonable directory and makes it easier to remember and tweak
the standard commands to deal with sessions.
More information in \tttt{:h sessions}.
One might use \tttt{:h views} to keep the state of one window,
but sessions keep all the states from all windows.
This entails a strategy to deal with Vim that is similar to
the use of
Byobu/Tmux/Screen\footnote{\url{https://en.wikipedia.org/wiki/Byobu_(software)},
\url{https://github.com/tmux/tmux/wiki},
\url{https://www.gnu.org/software/screen/}},
because one can rely on restoring the state of the windows.
The main limitation I found to this approach is that
Vim is not keeping track of the terminals opened.
If you open a terminal inside Vim with the \tttt{:term}
command, you will save the session as usual, but when loading
you get dummy empty windows for them and an error message.
More about Terminal-Job mode in Section~\ref{terminal}.

Autocommands are the standard way to define event-triggered routines in Vim.
These are often related to particular file types, but are also often
in defining the automated behavior of Vim.
If the autocommands are placed inside a configuration file (e.g. the
vimrc), the automated behavior is persistent across Vim instances.
E.g. in~\cite{vimrc} is found an autocommand for keeping track of the last inserted texts in the \tttt{".lkjh} registers (\tttt{@.lkjh} variables).

\subsection{Using Vim's modes}\label{modes}
These are the  basic and fully implemented modes of usage in Vim:
\begin{itemize}
  \item Normal mode: used for changing
  the position of the cursor or the text displayed
  at the window.
  A core goal of the normal mode it to support fast
  navigation of the document while allowing
  the typist to maintain the fingers on the home row
  (i.e. on the center of the keyboard).
  The mode is also used for manipulating text
  (e.g. copy, paste, delete, change case) and
    changing to other modes.
    More in Section~\ref{normal}.
  \item Insert mode: for inserting text. More in
    Section~\ref{sec:ins}.
  \item Command-line mode: for entering Ex commands. More in
    Section~\ref{sec:com}.
  \item Ex mode: similar to command-line mode,
  but more specialized for running various Ex commands.
    More in Section~\ref{sec:ex}.
  \item Visual mode: for making, manipulating and navigating
  selections of texts.
  \item Select mode: similar to visual mode but
  favors CUA\footnote{IBM Common User Access:
    \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
\end{itemize}

There is another basic mode, but it is not fully implemented:
the Terminal-Job mode (more in Section~\ref{terminal}.
There are seven additional modes which are mostly subordinate 
to the basic modes and that will be described when convenient.
The manual page for Vim modes can be accessed by typing
\texttt{:h vim-mode}.
Some of the modes are now further considered for
the achievement of
an overview of the Vim usage possibilities.

\subsubsection{Normal mode}\label{normal}
Sometimes also called navigation or command mode,
the normal mode is most powerful for
navigating, manipulating texts and changing to other modes.
The simplest of these three is changing to other modes:
type any of these letters to change to insert mode:
\texttt{iIaAoOsScC}. More on the transition between
normal and insert mode on Section~\ref{navIn}.
Type any of these characters to change to command-line mode:
\texttt{:/?}.
Type \texttt{Q} to enter Ex mode.
Type \texttt{v}, \texttt{V}, or \texttt{CTRL+V} to enter visual mode.

For very basic and naive navigation, one should check Section~\ref{minimum}.
There are many facilities to navigate Vim
as explained in \texttt{:h navigation}.
Most often, one uses:
\begin{itemize}
  \item \texttt{Ctrl+(d,u,f,b)} for half-page down and up
and whole page down and up, respectively, although these commands might
be set to scroll a different number of lines.
  \item \texttt{Ctrl+(e,y)} to move the window one line down or up.
  \item \texttt{(w,b,e)} to move to the next and previous word,
    and next end-of-word.
    There are motions to iterate over sequences of characters separated
    by special characters (e.g. punctuation and parenthesis) as
    specified by the output of \texttt{:se iskeyword}.
    To iterate over space-separated tokens, use \texttt{W,B,E}).
    To move to the end of last word, one might use \texttt{be}
    or \texttt{ge}.
The \texttt{),(} commands iterate through sentences,
		\tttt{\},\{} through text blocks separated by empty lines.
  \item \texttt{(fX,tX,FX, TX)} to move to or just before any X character,
	  \tttt{;} and \tttt{,} for next and previous found character.
  \item Search with \tttt{/} or \tttt{?}, although these are in truth command-line commands.
  \item \texttt{CTRL+(o,i)} to move to an older or newer position in
    jump list.
  \item \texttt{'X,`X} to move the cursor to a mark bind to the alphanumeric character X:
    \texttt{`X} moves to the exact position while \texttt{'X} moves to
    the first non-blank character of the line.
    A mark is registered by the user in any cursor position
    by typing \texttt{mX}, where X is any letter.
    If X is lowercase, the mark is local to the buffer (the file),
    if it is uppercase or numeric, it is global to the Vim session
    (cross buffers).
\end{itemize}

@@

There are way more commands to change the texts.
Some of them are discussed in Section~\ref{navIn}
because they involve a transition to the insert mode.
A thorough consideration of the commands in the normal mode
is found by executing \texttt{:h navigation}, 
\texttt{:h change.txt}, \tttt{:h index}.

\subsubsection{Insert mode}\label{sec:ins}
Once in the insert mode, the character keys
input the characters at the cursor position at the current buffer.
One can exit insert mode by pressing \texttt{<Esc>} (or \ttt{C-$[$} or
\ttt{C-C}),
and Vim will be put in the normal mode.
Most useful commands in insert mode include:
\begin{itemize}
  \item \ttt{C-O} to execute one and only command in normal mode.
	  This enters a secondary mode (see Section~\ref{modes})
  \item \ttt{C-R} to paste a register (a variable starting with
    '@', defined, copied or recorded through a \tttt{q} command in
    normal mode and as covered in Section~\ref{state}).
  \item \ttt{C-T} to indent current line.
  \item \ttt{C-U,W} to delete all chars from cursor to the beginning of
    the line or to the previous word.
  \item \ttt{C-N,P} to find next and previous keywords that match the prefix at hand.
  \item \ttt{C-X} commands for scrolling the window with multiple
    \ttt{C-E} and \ttt{C-Y} strokes and for some completion facilities.
\end{itemize}

@@

      \item Sessions, as described in Section~\ref{visual1}.

@@

        commands. They are very elementary use of registers which are executed over
        arbitrary but consistent textual patterns.

@@

      \item Run Python excerpts from a file being edited in an IPython shell. 

@@

        current highlighting scheme to a working Vimscript file

>>>>

\subsection{Further remarks about this document}\label{furRem}
This document is written is a DRY KISS
(Don't Repeat Yourself, Keep It Simple Stupid) style.
Complex is to master the use of Vim
and one finds sound
references in help files and a nice vimrc.
Therefore the following content is kept
as uncomplicated and original as possible.
Also, because of Vim's complexity and entailed
bond of this document to my usage,
there is an anthropological component
which is evident in the occasional use of the first person.
This can be understood as anthropological computer science~\cite{anPh,anPh2}
and considered to help
in the technological groundwork of the civil society.
Accordingly, here are some notes about my experience:
I've had experience with other editors, e.g. Kate, gedit, and Notepad2.
I used Vim for writing and editing computer code 
(Python, Javascript, C++, ChucK, bash, etc), markup-like content
(HTML, CSS, RDF, Markdown, \LaTeX, etc) and binary files.
Eventually, I edited other types of files, such as database files with
dozens of gigabytes.
Within Vim, I mostly write (web and scientific) software,
scientific articles, music, poems and short stories.

@@

\subsection{The bare minimum}\label{minimum}
You open a file at Vim startup by executing
the command: \texttt{\$ vim <filename>}.
Inside Vim, you start in the normal
mode, and might want to move around using
\texttt{h-j-k-l} for left-down-up-right.
To insert characters, move your
cursor to the desired location an press i,
which puts Vim in the insert mode.
Go back to normal mode by pressing
\texttt{<ESC>}, \ttt{C-$[$} or \texttt{<C-C>}.
You save the file by typing \texttt{:w<CR>},
and exit Vim by typing \texttt{:q<CR>}.
You can save and quit with \tttt{:wq<CR>}
or \tttt{:x<CR>} or \tttt{ZZ}.

\subsection{Vim help}
Help on using Vim is found in various places.
The standard resource is the Vim help files.
They are accessed by typing \texttt{:h <anything><CR>}
in normal mode.
Examples of such \texttt{<anything>} are:
color, navigation, :vs, vimtutor.
Type 
\texttt{:h usr\_toc<CR>}
to access the official User Manual,
which is considerably lengthy and complex
and is usually not read by users before they achieve a few years
of experience.
In learning Vim, one
might want to run the \texttt{\$ vimtutor} command
(outside Vim) to start the Vim Tutor.

There are good resources on the Web for learning
and tweaking Vim:
\begin{itemize}
  \item ``Vim Adventures'' is an online RPG game for practicing
  and memorizing Vim commands. This game is quite famous among Vim users.
  \item There are official and semi-official Vim sites e.g.:
  \url{www.vim.org}, \url{https://www.vi-improved.org} and
  \url{http://vim.wikia.com/}.
  \item Many hacks, understandings and general issues
  (e.g. how to make such a move) are asked and answered
  in online platforms (e.g. Quora, Stack Overflow, Stack Exchange, Reddit, Email list, IRC Channel).
  One often finds these links through a search engine.
  \item Many videos about Vim are publicly available.
  One might find them using a search engine
  (e.g. \url{http://derekwyatt.org/vim/tutorials/}, and
  \url{http://vimcasts.org}) or in Youtube and Vimeo.
\end{itemize}

% any other commands? files?
\subsection{Namespaces}\label{namespaces}
Vim is a text editor ouroboros~\cite{ouroWiki} because
text and writing alters text and writing.
This is achieved through a collection of namespaces where tokens have scalar
or arbitrarily complex values, or start automated routines.

\subsubsection{Commands and mappings}
There are commands, i.e. typing sequences which trigger actions,
for each mode:
\begin{itemize}
	\item in Normal mode all keys are mapped to commands.
		There is redundancy and additional commands
		using Ctrl and Shift keys.
		Some keys expect a second key,
    and have combinations not used (thus available for new mappings),
		specially the z and g keys.
    See Section~\ref{normal} and Appendix~\ref{notes}
    for more insights into the commands available in the
    normal mode.
	\item In the other modes, the sequences available for mappings are more obvious and abundant.
		One should look at \tttt{:h index} to know about all the standard mappings
		and use \tttt{:map} to list the user-defined mappings.
\end{itemize}

@@

A colon command can be written as a string
and executed by the \tttt{:execute} colon command.
E.g. \tttt{:execute 'vs afile.txt'} executes the \tttt{:vs afile.txt}
colon command..
As there are colon commands that execute commands in other
modes, e.g. \tttt{:normal ?\textasciicircum def},
the \tttt{:execute} is a way to build commands in any mode,
e.g. \tttt{:execute 'normal i'.string(atan(bufnr('\%'))).'<ESC>'}.

\subsubsection{Variables}
There are some types of variables in Vim:
\begin{itemize}
	\item Environment variables: names start with \$ and hold system
		variables, such as \tttt{\$PATH} and \tttt{\$PWD}.
	\item Option variables: names start with \& and are meant to control the behavior of the editor.
		One might change a value through set or let, e.g.
		\tttt{:set bg=light} or \tttt{:let \&bg=light}.
	\item Registers: start with @ and are meant for automation and transfer of texts (copy and paste).
    More on registers in Section~\ref{state} and scattered in this
    document.
  \item Internal variables are created with \tttt{:let} and preferably have a prefix:
	\tttt{b:}, \tttt{w:}, \tttt{t:}, \tttt{l:}, \tttt{s:},
		are local to the buffer, window, tab page, function, and
		sourced Vim file, respectively.
 \tttt{v:}, \tttt{g:} are globals, the first are predefined by Vim.
		\tttt{a:} is for function arguments.
		If there is no prefix, the variable is global or internal to a function if defined inside a function.
		More about internal variables in \tttt{:h internal-variables}.
	\item The value of a variable can be of the types: scalar, string, list, dictionary, function reference, etc (see \tttt{:h eval}).
\end{itemize}
You can \tttt{:echo} any of such variables or use them in expressions.
Notice that you will only be able to echo a \tttt{b:} variable inside
the buffer where it is defined.
For all the VimL capabilities, including loops, conditionals,
and builtin functions, refer to \tttt{:h vim-script} and Section~\ref{script}.
Classes are possible only in rudimentary forms, e.g. through dictionaries,
but the language is otherwise overall quite powerful,
specially in dealing with text and editor behavior, as expected.

\subsubsection{State lists}\label{state}
Vim keeps a number of useful lists which expresses the state of the editor:
\begin{itemize}
	\item All the entered commands are accessed through \tttt{:hist a}. 
    The tokens ``\tttt{a / e : i d}'' may be used for specific types of commands, such
		as search and colon commands.
	\item File buffers are kept with numeric ids. See buffers with \tttt{:ls} and load a buffer to the window with \tttt{:b <num or token in file name>}.
  \item The windows are listed in \tttt{:ls} with a character \tttt{a} in the second column, and are listed with \tttt{:tabs}.
	\item A tabs list can be reached through \tttt{:tabs}.
		It is usual both to show and hide the tabs bar (mapping
    in~\cite{vimrc}, discussion in Section~\ref{visual1}).
	\item Jumps are available through \tttt{:jumps}.
		One positions the cursor at each jump through \ttt{C-O} and \ttt{C-I}.
  \item Registers are available through the \tttt{:reg} command, as variables and through shortcuts in different modes.
    They also keep track or your copy, edition and deletion and are promptly defined
    by recording a typing sequence with the \tttt{q} normal command.
    Vim keeps only the last edition, in register \tttt{".}.
    An autocommand to keep the four latest inserts is in~\cite{vimrc}.
    A hack to keep the latest deletions and copies in the standard register
    might follow the same pattern, but use another \tttt{:h event} and
    monitor register \tttt{""} (maybe also monitor \tttt{"0}).
	\item An undo list is accessed through \tttt{:changes}.
	\item A list with all the sourced scripts in a Vim session is displayed through \tttt{:scriptnames}.
	\item The markers defined are listed with the \tttt{:marks} command.
		These are set by \tttt{mX} in normal mode, where X is the marker identifier.
		Uppercase letters are cross buffers.
	\item Quickfix and Location lists are populated through \tttt{:vim} and \tttt{:make}
		and variations, such as \tttt{:grep}.
		One might run \tttt{:vim /section/ \%} and then \tttt{:copen}
		to open the Quickfix window, where the lines of occurrence are in sequence
		and one can \ttt{CR} one of them to have the cursor in the main window active at
		the first character of the match.
		One might run \tttt{:lvim /section/ \%} and then \tttt{:lopen}
		to use the location-list window instead of the Quickfix, which
		is very similar, but one per window instead of one per buffer.
		More information in \tttt{:h quickfix}.
	\item A tags list have to be made in order to enable the use of tags.
		Most often one will generate the tags list using the exuberant ctags,
		which supports dozens of languages.
		E.g. \tttt{:!ctags-exuberant functions.py} or \tttt{!ctags-exuberant -R ./},
		and then using \ttt{C-]} to
		go to the position of tag under cursor, and \tttt{:vs tags} to open
		the tags file.
	\item The argument list holds a list of files to be edited or browsed. 
		The list can be input at Vim startup (e.g. \tttt{\$ vim file1.txt file2.py})
	or using commands (e.g. \tttt{:ar ./*}).
		The file being edited is changed by \tttt{:n} and \tttt{:p} commands,
		one might perform actions on each file in argument list using \tttt{:argdo}.
		All files in argument list are also in the buffers list.
    For further information, see \tttt{:h arglist} and Section~\ref{netrw}.
  \item All the autocommands are listed with \tttt{:au}.
    They are event-triggered actions Vim performs.
\end{itemize}

A window with information about the state of the editor
can be achieved through: \tttt{:source \$VIMRUNTIME/bugreport.vim}.
Also, this script might be examined because it has
a collection of commands to access various settings of Vim.
Another good list of commands to know about Vim's state is kept on
\url{http://vim.wikia.com/wiki/Displaying_the_current_Vim_environment}.
I would specially highlight the \tttt{:syntax} command because
it displays token groups and their meanings when run
inside e.g. a \tttt{.py}, \tttt{.vim} or help file.
More about using colors in Vim for syntax highlighting in Section~\ref{visual2}.

\subsubsection{On the persistence of visual cues about the editor state}\label{visual1}
You can keep track of the editor state though commands, as stated
in last subsection.
Also, one might rely on persistent visual
cues, specially the tabs bar, the status line, and the line
reserved for the command-line.
A good strategy I find is to have specialized visual cues of the state to make persistent or hide
and a mapping to toggle each of them.
Currently, I toggle byobu/screen/tmux bar with \ttt{F5},
status line and tabs bar with \tttt{<localleader>-T and B}, according to script~\cite{vimrc}.
I am mostly using the cleanest setting, toggling on the tabs bar and status line sometimes.
Numbering is always there, I rarely turn them off but keep the mappings \tttt{<leader>-n and N} to toggle just in case.
Highlight of last search string, and underline of unrecognized words (e.g. in English of Portuguese), are toggled with \tttt{<localleader>-l and p}.
Instead of keeping the status bar, I use \ttt{C-G} to know about the
file and \ttt{gC-G} to know more and rarely.
It seems not possible to remove the statusbar between horizontal splits.
After asking in online forums and experimenting, I realized that it seems
reasonable to keep at least one line dividing the windows, so
if it comes to it, I just \tttt{set statusline=-}.
Unfortunately, as far as I could dig, one will need to alter
the Vim's source code to enable a horizontal split without losing a line,
which, for me, seems far from the ideal.
An example solution is to implement a visual cue for the
first and/or last line of the windows in the line-numbers
column, or complete the spaces and empty chars with \$\$\$\$,

\subsubsection{On the persistence of the editor state}
For state persistence, one might keep an undo file for each file as
in~\cite{vimrc}.
Sessions are easy to manage, enabling one to save and load the
editor's state, with the opened windows, tabs, buffers, etc.
The mappings in~\cite{vimrc} keep the sessions
in a reasonable directory and makes it easy to remember and tweak
the standard commands to deal with sessions.
More information in \tttt{:h sessions}.
One might use \tttt{:h views} to keep the state of one window,
but sessions keep all the states from all windows.
This entails a strategy to deal with Vim that is similar to
the use of
Byobu/Tmux/Screen\footnote{\url{https://en.wikipedia.org/wiki/Byobu_(software)},
\url{https://github.com/tmux/tmux/wiki},
\url{https://www.gnu.org/software/screen/}},
because one can rely on restoring the state of the windows.
The main limitation I found to this approach is that
Vim is not keeping track of the terminals opened.
If you open a terminal inside Vim with the \tttt{:term}
command, you might save the session as usual, but when loading
you get a dummy empty window for the terminal and an error message.
More about Terminal-Job mode in Section~\ref{terminal}.

Autocommands are the standard way to define event-triggered routines in Vim.
These are often related to particular file types, but are also often
in defining some general automated behavior for Vim.
If the autocommands are placed inside a configuration file (e.g. the
vimrc), the automated behavior is persistent across Vim instances.
E.g. in~\cite{vimrc} is found an autocommand for keeping track of the last inserted texts in the \tttt{".lkjh} registers (\tttt{@.lkjh} variables).

\subsection{Using Vim's modes}\label{modes}
These are the  basic and fully implemented modes of usage in Vim:
\begin{itemize}
  \item Normal mode: used for changing
  the position of the cursor or the text displayed
  at the window.
  A core goal of the normal mode it to support fast
  navigation of the document while allowing
  the typist to maintain the fingers on the home row
  (i.e. on the center of the keyboard).
  The mode is also used for manipulating text
  (e.g. copy, paste, delete, change case) and
    changing to other modes.
    More in Section~\ref{normal}.
  \item Insert mode: for inserting text. More in
    Section~\ref{sec:ins}.
  \item Command-line mode: for entering Ex commands. More in
    Section~\ref{sec:com}.
  \item Ex mode: similar to the command-line mode,
  but more specialized for running various Ex commands.
    More in Section~\ref{sec:ex}.
  \item Visual mode: for making, manipulating and navigating
  selections of texts.
  \item Select mode: similar to visual mode but
  favors CUA\footnote{IBM Common User Access:
    \url{https://en.wikipedia.org/wiki/IBM_Common_User_Access}.}.
\end{itemize}

There is another basic mode, but it is not fully implemented:
the Terminal-Job mode (more in Section~\ref{terminal}).
There are seven additional modes which are mostly subordinate 
to the basic modes and that will be described when convenient.
The manual page for Vim modes can be accessed by typing
\texttt{:h vim-mode}.
Some of the modes are now further considered for
the achievement of
an overview of the Vim usage possibilities.

\subsubsection{Normal mode}\label{normal}
Sometimes also called navigation or command mode,
the normal mode is most powerful for
navigating, manipulating text and changing to other modes.
The simplest of these three is changing to other modes:
type any of these letters to change to insert mode:
\texttt{iIaAoOsScC}. More on the transition between
normal and insert mode on Section~\ref{navIn}.
Type any of these characters to change to command-line mode:
\texttt{:/?}.
Type \texttt{Q} to enter Ex mode.
Type \texttt{v}, \texttt{V}, or \texttt{CTRL+V} to enter visual mode.

For very basic and naive navigation, one should check Section~\ref{minimum}.
There are many facilities to navigate Vim
as explained in \texttt{:h navigation}.
Most often, one uses:
\begin{itemize}
  \item \texttt{Ctrl+(d,u,f,b)} for half-page down and up
and whole page down and up, respectively, although these commands might
be set to scroll a different number of lines.
  \item \texttt{Ctrl+(e,y)} to move the window one line down or up.
  \item \texttt{(w,b,e)} to move to the next and previous word,
    and next end-of-word.
    These are motions to iterate over sequences of characters separated
    by special characters (e.g. punctuation and parenthesis) as
    specified by the output of \texttt{:se iskeyword}.
    To iterate over space-separated tokens, use \texttt{W,B,E}).
    To move to the end of last word, one might use \texttt{be}
    or \texttt{ge}.
The \texttt{),(} commands iterate through sentences,
		\tttt{\},\{} through text blocks separated by empty lines.
    \tttt{[} and \tttt{]} are used in combination to browse sections.
  \item \texttt{(fX,tX,FX, TX)} to move, in the same line, to or just before any X character,
	  \tttt{;} and \tttt{,} for next and previous found character.
  \item Search with \tttt{/} or \tttt{?} (these are considered command-line commands).
  \item \texttt{CTRL+(o,i)} to move to an older or newer position in
    jump list.
  \item \texttt{'X,`X} to move the cursor to a mark bind to the alphanumeric character X.
    \texttt{`X} moves to the exact position while \texttt{'X} moves to
    the first non-blank character of the line.
    A mark is registered by the user in any cursor position
    by typing \texttt{mX}, where X is any letter.
    If X is lowercase, the mark is local to the buffer (the file),
    if it is uppercase or numeric, it is global to the Vim session
    (cross buffers).
\end{itemize}

@@

There are way more commands to change text.
Some of them are discussed in Section~\ref{navIn}
because they involve a transition to the insert mode.
A thorough consideration of the commands in the normal mode
is found by executing \texttt{:h navigation}, 
\texttt{:h change.txt}, \tttt{:h index}.

\subsubsection{Insert mode}\label{sec:ins}
Once in the insert mode, the character keys
input the characters at the cursor position in the current buffer.
One can exit insert mode by pressing \texttt{<Esc>} (or \ttt{C-$[$} or
\ttt{C-C}),
and Vim will be put in the normal mode.
Most useful commands in insert mode include:
\begin{itemize}
  \item \ttt{C-O} to execute one and only command in normal mode.
	  This enters a secondary mode (see Section~\ref{modes})
  \item \ttt{C-R} to paste a register (a variable starting with
    '@', defined, copied or recorded through a \tttt{q} command in
    normal mode and as covered in Section~\ref{state}).
  \item \ttt{C-T} to indent current line.
  \item \ttt{C-U,W} to delete all chars from cursor to the beginning of
    the line or to the previous word.
  \item \ttt{C-N,P} to find next and previous keywords that match the prefix at hand.
  \item \ttt{C-X} commands for scrolling the window with multiple
    \ttt{C-E} and \ttt{C-Y} strokes and for some completion facilities
    (\ttt{C-X}\ttt{C-(F,I,L,O,V} for file, identifier, whole line, omni and command-line completions). 
\end{itemize}

@@

      \item Sessions, as described in Section~\ref{visual1} and preliminarily implemented in~\cite{vimrc}.

@@

        commands. They comprise a very elementary use of registers which are executed over
        consistent textual patterns.

@@

      \item Run Python excerpts, from a file being edited, in an IPython shell. 

@@

        current highlighting scheme to a working VimL file
